---
output: 
  html_document:
title: "Reproducible Research: Peer Assessment 1"
---

```{r setup, include=FALSE}
# precondition: setwd() to the project directory
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data
```{r echo=TRUE}
library(ggplot2)
library(stringr)
if (!file.exists("./activity.zip")) {
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", 
                "./activity.zip", "auto")
}
if (!file.exists("./activity.csv")) {
  unzip("./activity.zip")
}

activity <- read.csv("activity.csv")

summary(activity)
```
## What is mean total number of steps taken per day?
First, calculating the total number of steps per day
```{r echo=TRUE}
total_steps <- aggregate(activity$steps,list(as.Date(activity$date)), function(x) sum(x, na.rm = TRUE))
names(total_steps) <- c("Date","Total.Steps")
ggplot(total_steps, aes(Date, Total.Steps), ylab="Steps") + geom_histogram(stat="identity") + scale_x_date()
```

Given those totals, we can arrive at the mean and median values
```{r echo=TRUE}
mean_steps <- mean(total_steps$Total.Steps, na.rm=TRUE)
median_steps <- median(total_steps$Total.Steps, na.rm=TRUE)
```

For this data set, the mean number of steps taken daily was `r mean_steps` per day and the median number of steps was `r median_steps` per day.

## What is the average daily activity pattern?
```{r echo=TRUE}
activity_intervals <- aggregate(activity$steps,list(str_pad(activity$interval, width=4,pad="0")), function(x) mean(x, na.rm=TRUE))
names(activity_intervals) <- c("Interval","Mean.Steps")
activity_intervals <- cbind(activity_intervals,Time=strptime(activity_intervals$Interval, format="%H%M"))
plot(data=activity_intervals, type="l", activity_intervals$Mean.Steps ~activity_intervals$Time, ylab="Mean Steps", xlab="Time of Day")
title("Activity by time of day")
abline(h=mean(activity_intervals$Mean.Steps), col="red")

peak <- activity_intervals[which(activity_intervals$Mean.Steps == max(activity_intervals$Mean.Steps)), ]$Time
```

From this data, it appears that the maximum activity occurs at `r strftime(peak,"%H:%M")`

## Imputing missing values
```{r echo=TRUE}
missing_values <- sum(is.na.data.frame(activity))
missing_steps <- sum(is.na.data.frame(activity$steps))
```
The dataset has `r missing_values` missing values. It appears that all of the missing values are in the steps count, with a total of `r missing_steps` missing values in that column.

To impute missing values, I considered using either the mean for the day, or the mean for the time interval. The problem with mean for the day is that there were entire days that were missing so I decided that the best approach was to use the mean for the time interval. 

```{r echo=TRUE}
activity$imputed_steps <- activity_intervals$Mean.Steps
activity$steps <- ifelse(is.na(activity$steps), activity$imputed_steps, activity$steps) 
```

With the imputed values, the graph looks like this now

```{r echo=TRUE}
total_steps_imputed <- aggregate(activity$steps,list(as.Date(activity$date)), function(x) sum(x, na.rm = TRUE))
names(total_steps_imputed) <- c("Date","Total.Steps")
ggplot(total_steps_imputed, aes(Date, Total.Steps), ylab="Steps") + geom_histogram(stat="identity") + scale_x_date()
mean_steps_imputed <- mean(total_steps_imputed$Total.Steps, na.rm=TRUE)
median_steps_imputed <- median(total_steps_imputed$Total.Steps, na.rm=TRUE)

```

With the imputed values, the mean steps per day is now `mean_steps_imputed` and the median is now `median_steps_imputed`

## Are there differences in activity patterns between weekdays and weekends?

We will now classify the dates as weekday or weekend and graph the activity levels using those categories.

```{r echo=TRUE}
activity$datetype <- ifelse(weekdays(as.Date(activity$date),abbreviate=TRUE) == "Sat" | weekdays(as.Date(activity$date),abbreviate=TRUE) == "Sun", "weekend","weekday")
activity$datetype <- as.factor(activity$datetype)
```

The final task is to create a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days. 

```{r echo=TRUE}
mean_steps <- aggregate(activity$steps,list(str_pad(activity$interval, width=4,pad="0"), activity$datetype), function(steps) mean(steps, na.rm=TRUE))
names(mean_steps) <- c("Interval","Day.Type","Mean.Steps")
mean_steps <- cbind(mean_steps,Time=strptime(mean_steps$Interval, format="%H%M"))
ggplot(mean_steps, aes(Time, Mean.Steps)) + geom_line() + facet_grid(Day.Type ~ .) + labs(y="Number of steps", x="Interval")

```